\msTitle

### 2014-02-19 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
Updating formating of manuscripts.
Updating preamble to remove section title and caption formatting
options. Using `\maketitle` in dpp manuscript to create title page.
Changing all `section*` calls to `section` in dpp manuscript.

\section*{Introduction}

@@

\section*{Methods}

\subsection*{The model}
In this section, we describe our model, which is a modification of the model
implemented in \msb \citep{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{https://github.com/joaks1/dpp-msbayes}).
Our analyses described below utilize the freely avaliable, open-source software
package \pymsbayes (\href{https://github.com/joaks1/PyMsBayes}{https://github.com/joaks1/PyMsBayes}),
which is a Python API that provides a multi-processing interface to \msb and
\dppmsbayes.
We performed the work described below following the principles of
Open Notebook Science.
Using version-control software, we make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{https://github.com/joaks1/msbayes-experiments}.
All information necessary to reproduce our results is provided there.
We follow much of the notation of \citet{Oaks2012}, but modify it to
aid in our description of the new model.
A summary of our notation can be found in Table \ref{tabNotation}.

@@

\subsubsection*{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The likelihood terms of Equation \ref{eq:fullModelCompact} can
be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$,
is the probability of the sequence alignment of locus $j$ for population pair
$i$ given the gene tree and \hky substitution model parameters \citep[i.e., the
``Felsenstein likelihood'';][]{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams{} contains the following demographic parameters: The
mutation-scaled effective sizes ($\myTheta{} = 4N\mutationRate$) of the
ancestral, \ancestralTheta{}, and descendant populations, \descendantTheta{1}{}
and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persist during bottlenecks that begin immediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \citep{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

\subsubsection*{Prior terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape-parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate scaling parameters
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identically
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be
non-zero, the prior term $p(\recombinationRate)$ would be added
to Equation \ref{eq:fullModelCompact}, and the prior would be
$\recombinationRate \sim Gamma(\gammaShape{\recombinationRate},
\gammaScale{\recombinationRate})$.
\end{linenomath}

@@

\subsubsection*{Priors on divergence models}
\begin{linenomath}
The prior term for the vector of divergence times for each of
the \npairs{} pairs of populations can be expanded as
\begin{equation}
    p(\divTimeMapVector) = p(\divTimeIndexVector)p(\divTimeVector \given \divTimeIndexVector),
    \label{eq:divModelPrior}
\end{equation}
where \divTimeVector is an ordered set of unique divergence-time parameters
$\{\divTime{1}, \ldots, \divTime{\divTimeNum}\}$ whose length
\divTimeNum can range from 1 to \npairs{},
and \divTimeIndexVector is the vector of indices of divergence-time parameters 
$(\divTimeIndex{1}, \ldots, \divTimeIndex{\npairs{}})$
that map the divergence times in \divTimeVector to each of the \npairs{}
pairs of populations.
Thus, \divTimeMapVector is the result of applying the mapping function
\begin{equation}
    f(\divTimeVector, \divTimeIndexVector, i) = \divTime{\divTimeIndex{i}}
    \label{eq:divTimeMapFunction}
\end{equation}
to each population pair $i$, such that
$\divTimeMapVector = (\divTimeMap{1} = f(\divTimeVector, \divTimeIndexVector,
1), \ldots, \divTimeMap{\npairs{}} = f(\divTimeVector, \divTimeIndexVector,
\npairs{}))$.
\end{linenomath}

@@

\subsection*{Differences between our model and the original \msb model}
\subsubsection*{The prior on divergence models}
One of the key differences between our model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
We denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, we place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
We introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

\subsubsection*{Estimating ordered divergence models}
As mentioned above, \msb samples over unordered divergence models
(i.e., unordered partitions of the \npairs{} pairs of populations).
That is, the identity of each population pair, and all the information
associated with it, is discarded.
In our implementation, inference can be done on either unordered or ordered
models of divergence.
This is discussed in more detail in the description of the ABC implementation
below.

\subsubsection*{The priors on nuisance parameters}
We have replaced the use of continuous uniform distributions for priors on many
of the model's parameters (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \bottleScalar{1}{},
\bottleScalar{2}{}, \recombinationRate, \migrationRate{}) with more flexible
parametric distributions from the exponential family.
We introduce gamma-distributed priors for rate parameters that have a sample
space of all positive real numbers (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \recombinationRate,
\migrationRate{}), and beta-distributed priors for parameters that are
proportions bounded by zero and one (\bottleScalar{1}{} and
\bottleScalar{2}{}).
These priors provide an investigator with much greater flexibility in
expressing prior uncertainty regarding the parameters of the model.

@@

\subsubsection*{Flexibility in parameterizing the model}
In our implementation, we provide the ability to control the richness of the
model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.

@@

\subsubsection*{Time scale}
As described above, in our model divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by \citet{Oaks2012}, in \msb, \myTheta{C} is half of the upper
limit of the continuous uniform prior on the mean of the descendant population
sizes.

\subsection*{ABC estimation of the posterior of the model}
\subsubsection*{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameter values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which we denote \hpvector{},
from their respective prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments;
% (these are the same statistics
% calculated from the observed alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statistic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, we obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.

@@

\subsubsection*{Ordering of taxon-specific summary statistics}
As alluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, the summary statistics are re-ordered by descending values of average
pairwise differences between the descendant populations
\citep[$\pi_b$;][]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of discrete
divergence models \divTimeIndexVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

\subsubsection*{Obtaining an approximate posterior from the prior samples}
We use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and the
difference is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the desired number of
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \citep{Beaumont2002,Blum2009,Leuenberger2010}.
For our analyses, we use the general linear model (GLM) regression adjustment
\citep{Leuenberger2010} as implemented in \abctoolbox
\citep[v1.1;][]{ABCtoolbox}, which \citet{Oaks2012} showed performed very
similarly to weighted local-linear regression and multinomial logistic
regression adjustments \citep{Beaumont2002} for \msb posteriors.

>>>>

\section{Introduction}

@@

\section{Methods}

\subsection{The model}
In this section, we describe our model, which is a modification of the model
implemented in \msb \citep{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{https://github.com/joaks1/dpp-msbayes}).
Our analyses described below utilize the freely avaliable, open-source software
package \pymsbayes (\href{https://github.com/joaks1/PyMsBayes}{https://github.com/joaks1/PyMsBayes}),
which is a Python API that provides a multi-processing interface to \msb and
\dppmsbayes.
We performed the work described below following the principles of
Open Notebook Science.
Using version-control software, we make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{https://github.com/joaks1/msbayes-experiments}.
All information necessary to reproduce our results is provided there.
We follow much of the notation of \citet{Oaks2012}, but modify it to
aid in our description of the new model.
A summary of our notation can be found in Table \ref{tabNotation}.

@@

\subsubsection{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The likelihood terms of Equation \ref{eq:fullModelCompact} can
be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$,
is the probability of the sequence alignment of locus $j$ for population pair
$i$ given the gene tree and \hky substitution model parameters \citep[i.e., the
``Felsenstein likelihood'';][]{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams{} contains the following demographic parameters: The
mutation-scaled effective sizes ($\myTheta{} = 4N\mutationRate$) of the
ancestral, \ancestralTheta{}, and descendant populations, \descendantTheta{1}{}
and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persist during bottlenecks that begin immediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \citep{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

\subsubsection{Prior terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape-parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate scaling parameters
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identically
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be
non-zero, the prior term $p(\recombinationRate)$ would be added
to Equation \ref{eq:fullModelCompact}, and the prior would be
$\recombinationRate \sim Gamma(\gammaShape{\recombinationRate},
\gammaScale{\recombinationRate})$.
\end{linenomath}

@@

\subsubsection{Priors on divergence models}
\begin{linenomath}
The prior term for the vector of divergence times for each of
the \npairs{} pairs of populations can be expanded as
\begin{equation}
    p(\divTimeMapVector) = p(\divTimeIndexVector)p(\divTimeVector \given \divTimeIndexVector),
    \label{eq:divModelPrior}
\end{equation}
where \divTimeVector is an ordered set of unique divergence-time parameters
$\{\divTime{1}, \ldots, \divTime{\divTimeNum}\}$ whose length
\divTimeNum can range from 1 to \npairs{},
and \divTimeIndexVector is the vector of indices of divergence-time parameters 
$(\divTimeIndex{1}, \ldots, \divTimeIndex{\npairs{}})$
that map the divergence times in \divTimeVector to each of the \npairs{}
pairs of populations.
Thus, \divTimeMapVector is the result of applying the mapping function
\begin{equation}
    f(\divTimeVector, \divTimeIndexVector, i) = \divTime{\divTimeIndex{i}}
    \label{eq:divTimeMapFunction}
\end{equation}
to each population pair $i$, such that
$\divTimeMapVector = (\divTimeMap{1} = f(\divTimeVector, \divTimeIndexVector,
1), \ldots, \divTimeMap{\npairs{}} = f(\divTimeVector, \divTimeIndexVector,
\npairs{}))$.
\end{linenomath}

@@

\subsection{Differences between our model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between our model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
We denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, we place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
We introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

\subsubsection{Estimating ordered divergence models}
As mentioned above, \msb samples over unordered divergence models
(i.e., unordered partitions of the \npairs{} pairs of populations).
That is, the identity of each population pair, and all the information
associated with it, is discarded.
In our implementation, inference can be done on either unordered or ordered
models of divergence.
This is discussed in more detail in the description of the ABC implementation
below.

\subsubsection{The priors on nuisance parameters}
We have replaced the use of continuous uniform distributions for priors on many
of the model's parameters (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \bottleScalar{1}{},
\bottleScalar{2}{}, \recombinationRate, \migrationRate{}) with more flexible
parametric distributions from the exponential family.
We introduce gamma-distributed priors for rate parameters that have a sample
space of all positive real numbers (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \recombinationRate,
\migrationRate{}), and beta-distributed priors for parameters that are
proportions bounded by zero and one (\bottleScalar{1}{} and
\bottleScalar{2}{}).
These priors provide an investigator with much greater flexibility in
expressing prior uncertainty regarding the parameters of the model.

@@

\subsubsection{Flexibility in parameterizing the model}
In our implementation, we provide the ability to control the richness of the
model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.

@@

\subsubsection{Time scale}
As described above, in our model divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by \citet{Oaks2012}, in \msb, \myTheta{C} is half of the upper
limit of the continuous uniform prior on the mean of the descendant population
sizes.

\subsection{ABC estimation of the posterior of the model}
\subsubsection{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameter values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which we denote \hpvector{},
from their respective prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments;
% (these are the same statistics
% calculated from the observed alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statistic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, we obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.

@@

\subsubsection{Ordering of taxon-specific summary statistics}
As alluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, the summary statistics are re-ordered by descending values of average
pairwise differences between the descendant populations
\citep[$\pi_b$;][]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of discrete
divergence models \divTimeIndexVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

\subsubsection{Obtaining an approximate posterior from the prior samples}
We use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and the
difference is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the desired number of
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \citep{Beaumont2002,Blum2009,Leuenberger2010}.
For our analyses, we use the general linear model (GLM) regression adjustment
\citep{Leuenberger2010} as implemented in \abctoolbox
\citep[v1.1;][]{ABCtoolbox}, which \citet{Oaks2012} showed performed very
similarly to weighted local-linear regression and multinomial logistic
regression adjustments \citep{Beaumont2002} for \msb posteriors.

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \citet{Oaks2012}, we characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
We selected four model priors for these analyses (Table~\ref{tabPriors}):
(1) A model to represent the original \msb model, \modelOld, with
priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim U(0,10)$,
$\ancestralTheta{} \sim U(0, 0.05)$,
and
$\descendantThetaMean{} \sim U(0, 0.05)$;
(2) a model with the \msb prior on unordered divergence models, but with
exponential priors on nuisance parameters,
\modelUshaped, with priors
$\divTimeIndexVector \sim \priorOld$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
and
$\dThetas \sim Exp(mean=0.025)$.
(3) a model with uniform prior probability over unordered divergence models,
\modelUniform, with priors
$\divTimeIndexVector \sim \priorUniform$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
and
$\dThetas \sim Exp(mean=0.025)$;
and
(4) a Dirichlet-process prior model, \modelDPP, with priors
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(2,2)}$,
$\divTime{} \sim Exp(mean=2.887)$,
$\ancestralTheta{} \sim Exp(mean=0.025)$,
and
$\dThetas \sim Exp(mean=0.025)$.
We selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
We selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination,
and unordered divergence models.
For all of our simulations, we used a simulated data structure of eight
population pairs, with a single 1000 base-pair locus sampled from 10
individuals from each population.
